### 2025-06-18 manuscript/main.tex
minprob, BNF subpar and semanticsModel

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage{multicol}
\usepackage{fontspec}

@@

\usepackage{microtype}
\usepackage{graphicx}
\usepackage[colorlinks,linkcolor=blue,urlcolor=cyan]{hyperref}
\usepackage{titlesec}
\titleformat{\chapter}[hang]{\normalfont\huge\bfseries}{}{0pt}{}
%=== Metadata ===
\newcommand{\thesistitle}{Your Thesis Title Here}
\newcommand{\authorname}{Rafael Tanzer}
\newcommand{\supervisor}{Michele Chiaria, Simone , Francesco Pontigia and Ezzio Bartocci}
\newcommand{\institution}{TU Vienna}
\newcommand{\department}{Cyber-Physical-Systems, ..., ...}
\newcommand{\logoimage}{graphics/logo.png} % Path to your logo image
\newcommand{\submissionmonth}{May}
\newcommand{\submissionyear}{2025}
%================

@@

\subsection{Backus–Naur Form (BNF) and Variants}
Even though production rules, alphabet specifications, and automaton definitions suffice to unambiguously define the set of valid strings/sentences in a language, their notation
tends to be opaque and cumbersome in practice, prompting interest in forming intuitively comprehensible grammar definitions. The Backus-Naur Form (BNF),
created by John Backus with contributions by Peter Naur and released in their Algol-60 report. \cite{ALGOL60}, was a pivotal introduction furnishing a concise, human-readable syntax for language generation.
By expressing each rule as
\[
  \langle\mathrm{nonterminal}\rangle \;::=\; expansion_1 \;|\; expansion_2 \;|\;\dots
\]
BNF allows both sequencing and alternation of multiple expansions, enabling language designers to articulate complex, nested structures without exposing the underlying
automaton states or transition tables. This declarative approach not only supports rigorous specification but also facilitates mechanical parsing, thereby advancing the practice of compiler development and language tooling.

@@

??reference to BNF of MiniPRob from BNF section??
MiniProb’s syntax is carefully designed to comply with the constraints of \textbf{Operator Precedence Languages (OPLs)}, ensuring compatibility with \textbf{Operator Precedence Automata (OPAs)} and their probabilistic extensions. To achieve this, the language avoids syntactic ambiguity and enforces structured, well-bracketed constructs that can be parsed deterministically using an \textbf{Operator Precedence Matrix (OPM)}. Within these constraints, MiniProb adopts a \textit{procedural and recursive} programming model, supporting user-defined functions with call-return semantics, local variable declarations, and block-structured control flow. Its syntax includes familiar C-like constructs such as \code{if}/\code{else}, \code{while}, and \code{try}/\code{catch}, alongside typed variables (e.g., \code{bool}, \code{s8}, \code{u16}, including fixed-size arrays). Extending this foundation, MiniProb incorporates core primitives from \textit{probabilistic programming languages}: \code{observe} statements condition execution on logical expressions (with trace rejection semantics), \code{query} triggers inference over program executions, and distributions such as \code{Bernoulli} and \code{Uniform} enable stochastic behavior. Together, these features allow MiniProb programs to be translated into \textbf{probabilistic OPAs (pOPAs)}, supporting formal verification and probabilistic reasoning over recursive computations.

@@

>>>>

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage{minted}
\usepackage{multicol}
\usepackage{fontspec}

@@

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage[colorlinks,linkcolor=blue,urlcolor=cyan]{hyperref}
\usepackage{titlesec}
\titleformat{\chapter}[hang]{\normalfont\huge\bfseries}{}{0pt}{}
%=== Metadata ===
\newcommand{\thesistitle}{Your Thesis Title Here}
\newcommand{\authorname}{Rafael Tanzer}
\newcommand{\supervisor}{Michele Chiaria, Simone , Francesco Pontigia and Ezzio Bartocci}
\newcommand{\institution}{TU Vienna}
\newcommand{\department}{Cyber-Physical-Systems, ..., ...}
\newcommand{\logoimage}{graphics/logo.png} % Path to your logo image
\newcommand{\submissionmonth}{May}
\newcommand{\submissionyear}{2025}
%================

@@

\subsection{Backus–Naur Form (BNF) and Variants}
Even though production rules, alphabet specifications, and automaton definitions suffice to unambiguously define the set of valid strings/sentences in a language, their notation
tends to be opaque and cumbersome in practice, prompting interest in forming intuitively comprehensible grammar definitions. The Backus-Naur Form (BNF),
created by John Backus with contributions by Peter Naur and released in their Algol-60 report. \cite{ALGOL60}, was a pivotal introduction furnishing a concise, human-readable syntax for language generation.
By expressing each rule as
\[
  \langle\mathrm{nonterminal}\rangle \;::=\; expansion_1 \;|\; expansion_2 \;|\;\dots
\]
BNF allows both sequencing and alternation of multiple expansions, enabling language designers to articulate complex, nested structures without exposing the underlying
automaton states or transition tables. This declarative approach not only supports rigorous specification but also facilitates mechanical parsing, thereby advancing the practice of compiler development and language tooling.
??Describe constraints of BNF??

@@

\begin{wrapfigure}{l}{0.5\textwidth}
  \centering
  \[
    \begin{array}{rcl}
      prog
       & \to  & (decl\;\texttt{;})^* \;func^+                                                     \\[6pt]

      decl
       & \to  & type\;\mathit{id}\;(\texttt{,}\;\mathit{id})^*                                    \\[6pt]

      func
       & \to  & \mathit{id}\;\texttt{(}
      (\,type\;\texttt{\&}?\;\mathit{id}
      (\texttt{,}\;type\;\texttt{\&}?\;\mathit{id})^*\,)?
      \texttt{)}                                                                                  \\[-2pt]
       &      & \texttt{\{}(decl\;\texttt{;})^*\;stmt^+\texttt{\}}                                \\[6pt]

      fcall
       & \to  & \mathit{id}\;\texttt{(}
      [\,e\;(\texttt{,}\;e)^*\,]
      \texttt{)}                                                                                  \\[6pt]

      stmt
       & \to  & lval\;\texttt{=}\;e\;\texttt{;}                                                   \\[-2pt]
       & \mid & lval\;\texttt{=}\;\texttt{Distribution(...)}\;\texttt{;}                          \\[-2pt]
       & \mid & (\texttt{query})?\;fcall\;\texttt{;}                                              \\[-2pt]
       & \mid & \texttt{observe}\;e\;\texttt{;}                                                   \\[-2pt]
       & \mid & \texttt{throw}\;\texttt{;}                                                        \\[-2pt]
       & \mid & \texttt{if}\;\texttt{(}e\texttt{)}\;\{\;stmt^*\;\}\;\texttt{else}\;\{\;stmt^*\;\} \\[-2pt]
       & \mid & \texttt{while}\;\texttt{(}e\texttt{)}\;\{\;stmt^*\;\}                             \\[-2pt]
       & \mid & \texttt{try}\;\{\;stmt^*\;\}\;\texttt{catch}\;\{\;stmt^*\;\}                      \\[6pt]

      lval
       & \to  & \mathit{id}\;(\texttt{[}e\texttt{]})?                                             \\[6pt]

      type
       & \to  & \texttt{bool}
      \;\mid\;(s\mid u)\;int\;(\texttt{[}int\texttt{]})?                                          \\[6pt]
    \end{array}
  \]
  \caption{Compacted grammar of \texttt{MiniProb}.}
  \label{fig:ebnf_grammar}
\end{wrapfigure}

??reference to BNF of MiniPRob from BNF section??

A compacted grammar for MiniProb is shown in \ref{fig:ebnf_grammar} containing the core elements providing enough understanding?>.
A full BNF grammar can be found in the appendix see \ref{fig:bnf_grammar}.
A \emph{prog} consists of zero or more top-level declarations followed by one or more function definitions. Each \emph{decl} introduces one or more variables of a
given \emph{type}, which may be \code{bool}, signed (\code{s}) or unsigned (\code{u}) integers of certain bit-width (e.g., \code{s8}, \code{u16}),
and optionally fixed-size arrays (e.g., \code{u8[10u4]}). Functions (\emph{func}) consist of an identifier, an optional list of typed parameters (passed by value or
reference via \code{\&}), and a block containing optional local declarations and at least one statement.

Within function bodies, MiniProb allows for local declarations and familiar C-like statements, including variable assignments (\code{lval = \emph{e};}),
function calls, and control flow structures such as \code{if}/\code{else}, \code{while} loops, and structured exception handling via \code{try}/\code{catch},
with \code{throw;} used to abort execution. Left-hand values (\emph{lval}) may refer to scalar variables or indexed array elements, while types follow a compact
convention, consisting of Booleans and signed or unsigned integers of fixed width.

Probabilistic constructs are central to MiniProb's design. These include sampling from distributions via statements like \code{lval = Distribution(...);},
currently supporting both \code{Bernoulli} and \code{Uniform} distributions. The \code{observe \emph{e};} construct allows conditioning execution on logical expressions,
rejecting traces that do not satisfy the observed condition. Inference is triggered through the \code{query \emph{fcall};} statement, which indicates
the probability evaluation of a function's outcome given a set of observations.

In addition to structured control, MiniProb supports expressions built from literals, variables, and function calls, composed using arithmetic
operations (\code{+}, \code{-}, \code{*}, \code{/}, \code{\%}), comparison operators (\code{==}, \code{!=}, \code{<}, \code{<=}, etc.), and Boolean connectives
(\code{!}, \code{\&\&}, \code{||}). These expressions can appear on the right-hand side of assignments or as arguments.

Beyond the standards, MiniProb introduces \emph{probabilistic assignments} to express uncertain choices: $lval = e_1 \{ e_2 : e_3 \} \; e_4$.
This construct selects $e_1$ with probability $p = e_2 / e_3$, and $e_4$ with probability $1 - p$.
Multiple such choice blocks can be chained to define multi-way probabilistic decisions, enabling a compact encoding of stochastic control.

Identifiers follow the pattern \code{[a-zA-Z\_][a-zA-Z0-9\_\.:$\sim$]*}, allowing names with symbols like underscores, colons, dots, and tildes.
Integer literals are annotated explicitly with both value and type, e.g., \code{42s8}, \code{7u16}, where the suffix denotes signedness and bit width.
Boolean literals are written plainly as \code{true} and \code{false}.

@@

\begin{figure}[htbp]
  \centering
  \begin{minipage}[t]{0.48\textwidth}
    \[
      \begin{array}{rcl}
        prog
         & \to  & (decl\;\texttt{;})^*\;\;func\;func\_list?                                                                \\[6pt]

        decl\_list
         & \to  & decl\;\texttt{;}\;decl\_list
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        func\_list
         & \to  & func\;func\_list
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        decl
         & \to  & type\;\mathit{id}\;id\_list                                                                              \\[6pt]

        id\_list
         & \to  & \texttt{,}\;\mathit{id}\;id\_list
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        func
         & \to  & \mathit{id}\;\texttt{(}
        [\,type\;\texttt{\&}?\;\mathit{id}
          (\texttt{,}\;type\;\texttt{\&}?\;\mathit{id})^*\,]?
        \texttt{)}                                                                                                         \\
         &      & \texttt{\{}(decl\;\texttt{;})^*\;block\;\texttt{\}}                                                      \\[6pt]

        block
         & \to  & stmt\;stmt\_list                                                                                         \\[6pt]

        block\_opt
         & \to  & block
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        stmt\_list
         & \to  & stmt\;stmt\_list
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        stmt
         & \to  & lval\;\texttt{=}\;expr\;\texttt{;}
        \;\mid\; lval\;\texttt{=}\;\texttt{Distribution(...)}\;\texttt{;}                                                  \\
         & \mid & \texttt{query}\;func\_call\;\texttt{;}
        \;\mid\; \texttt{observe}\;expr\;\texttt{;}                                                                        \\
         & \mid & func\_call\;\texttt{;}
        \;\mid\; \texttt{throw}\;\texttt{;}                                                                                \\
         & \mid & \texttt{if}\;\texttt{(}expr\texttt{)}\;\{\;stmt^*\;\}\;\texttt{else}\;\{\;stmt^*\;\}                     \\
         & \mid & \texttt{while}\;\texttt{(}expr\texttt{)}\;\{\;stmt^*\;\}                                                 \\
         & \mid & \texttt{try}\;\{\;stmt^*\;\}\;\texttt{catch}\;\{\;stmt^*\;\}                                             \\[6pt]

        assignment
         & \to  & lval\;\texttt{=}\;assign\_rhs                                                                            \\[6pt]

        assign\_rhs
         & \to  & expr
        \;\mid\; distribution                                                                                              \\[6pt]

        func\_call
         & \to  & \mathit{id}\;\texttt{(}
        (expr\;(\texttt{,}\;expr)^*)?
        \texttt{)}                                                                                                         \\[6pt]

        arg\_list\_opt
         & \to  & arg\_list
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        arg\_list
         & \to  & arg\;arg\_list\_tail                                                                                     \\[6pt]

        arg\_list\_tail
         & \to  & \texttt{,}\;arg\;arg\_list\_tail
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        arg
         & \to  & expr                                                                                                     \\[6pt]

        lval
         & \to  & \mathit{id}
        \;\mid\; \mathit{id}\;\texttt{[}expr\texttt{]}                                                                     \\[6pt]

        distribution
         & \to  & \texttt{Bernoulli(}expr\texttt{,}expr\texttt{)}
        \;\mid\; \texttt{Uniform(}expr\texttt{,}expr\texttt{)}                                                             \\[6pt]

        param
         & \to  & type\;\texttt{\&}?\;\mathit{id}                                                                          \\[6pt]

        param\_list
         & \to  & param\;param\_list\_tail                                                                                 \\[6pt]

        param\_list\_tail
         & \to  & \texttt{,}\;param\;param\_list\_tail
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        expr
         & \to  & logical\_or\;prob\_assign\_opt                                                                           \\[6pt]

        prob\_assign\_opt
         & \to  & prob\_assign
        \;\mid\; \epsilon                                                                                                  \\[6pt]

        prob\_assign
         & \to  & prob\_choice\;expr                                                                                       \\[6pt]

        prob\_choice
         & \to  & \texttt{\{}\;expr\;\texttt{:}\;expr\;\texttt{\}}                                                         \\[6pt]

        logical\_or
         & \to  & logical\_and\;(\texttt{||}\;logical\_and)^*                                                              \\[6pt]

        logical\_and
         & \to  & comparison\;(\texttt{\&\&}\;comparison)^*                                                                \\[6pt]

        comparison
         & \to  & term\;(comparison\_op\;term)^*                                                                           \\[6pt]

        comparison\_op
         & \to  & \texttt{>}\;\mid\;\texttt{>=}\;\mid\;\texttt{<}\;\mid\;\texttt{<=}\;\mid\;\texttt{==}\;\mid\;\texttt{!=} \\[6pt]
      \end{array}
    \]
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.48\textwidth}
    \[
      \begin{array}{rcl}

        term
         & \to & factor\;(\texttt{+}\;\mid\;\texttt{-})\;factor^*                   \\[6pt]

        factor
         & \to & division\;(\texttt{*}\;division)^*                                 \\[6pt]

        division
         & \to & unary\;((\texttt{/}\;\mid\;\texttt{\%})\;unary)^*                  \\[6pt]

        unary
         & \to & (\texttt{!}\;\mid\;\texttt{-})\;unary
        \;\mid\; primary                                                            \\[6pt]

        primary
         & \to & bool
        \;\mid\; int\_literal
        \;\mid\; lval
        \;\mid\; \texttt{(}expr\texttt{)}                                           \\[6pt]

        identifier
         & \to & letter\;ident\_tail                                                \\[6pt]

        ident\_tail
         & \to & letter\_or\_digit\;ident\_tail
        \;\mid\; \epsilon                                                           \\[6pt]

        letter
         & \to & \texttt{a} \;\mid\; \dots \;\mid\; \texttt{Z} \;\mid\; \texttt{\_} \\[6pt]

        letter\_or\_digit
         & \to & letter \;\mid\; digit                                              \\[6pt]

        digit
         & \to & \texttt{0} \;\mid\; \dots \;\mid\; \texttt{9}                      \\[6pt]

        type
         & \to & \texttt{bool}
        \;\mid\; int\_spec
        \;\mid\; int\_spec\;\texttt{[}int\_literal\texttt{]}                        \\[6pt]

        digit\_seq
         & \to & digit\;digit\_seq
        \;\mid\; digit                                                              \\[6pt]

        int\_spec
         & \to & \texttt{u}\;digit\_seq
        \;\mid\; \texttt{s}\;digit\_seq                                             \\[6pt]

        int\_literal
         & \to & digit\_seq\;int\_spec                                              \\[6pt]

        bool
         & \to & \texttt{true} \;\mid\; \texttt{false}
      \end{array}
    \]
  \end{minipage}
  \caption{Full grammar of MiniProb DSL}
  \label{fig:bnf_grammar}
\end{figure}
